Estimating the replicability of psychology experiments after an initial failure to replicate

### 2024-10-14 manuscript/manuscript.Rmd
second proofreading pass

In psychology, large scale replication projects have found that around half of studies successfully replicate.
Across 100 studies with positive results, the reproducibility project in psychology (RP:P) replicated 36%-47% depending on the metric for replication success [@opensciencecollaboration].
With large multi-site samples, Many Labs 1 replicated 11/13 effects (84.6%), Many Labs 2 replicated 14/28 effects (50%), and Many Labs 3 replicated 3/10 effects (30%) [@klein2014; @klein2018; @ebersole2016].
@camerer2018 examined 21 behavioral social science studies and successfully replicated 12-14 of them (57%-67%) depending on the metric used.
@boyce2023 reported an average subjective replication score of 49% for 176 replications primary in psychology.

Psychology is not the only discipline where many studies do not replicate.
Large scale replication projects in other disciplines have found 39/97 of studies with positive effects (40%) replicating in cancer biology [@errington2021], 11/18 studies (61%) replicating in economics [@camerer2016], and 31/40 studies (78%) replicating in experimental philosophy [@cova2021].
Across these disciplines, and especially in psychology, large-scale replications indicate that the probability of scientists failing to replicate published findings is substantial.

@@

Other reasons for replication failure are primarily due to the design of the replication study.
A replication could observe different results from the original due to methodological differences.
These differences include intentional adaptations, unavoidable changes from lack of access to the original instructions or materials, or changes that the replicators did not expect to be critical.
A difference in the details of recruitment, such as a change from recruiting participants in person to online, might mean that particular details of a study implementation would no longer be appropriate, or the data quality controls might be inadequate.
If there are differences in time, place, or subject population between the two studies, researchers may need to change the materials, instructions, or procedure to adapt the experimental paradigm to the new context.
Such adaptation creates more challenges in interpretation: too many changes could cause differing results, but so could too few changes.
In sum, when replications fail, we generally do not know why they failed, although we may speculate.

@@

While it is possible to speculate after the fact about potential causes of a single replication failure, in most circumstances these causes are hard to discern (absent clear evidence of errors).
Re-replication can serve as a way to test explanations for replication failure.
Re-replications can make making modifications to the first replication and give another chance for researchers to measure an effect of interest.
But what are the chances that a re-replication succeeds?

@@

While @ebersole2020 looked at re-replication success rate in a high resource setting with access to large, multi-site samples and expert input, most early career scientists do not have comparable resources to devote to re-replication attempts.
On the other hand, early career researchers may also be prone to a different distribution of causes of initial replication failure.
Thus, it is an open question what the re-replication rate is for failed replications by early career researchers.
Further, estimating this rate could inform researchers' decisions on what to do next after failed replications.

@@

PSYCH 251 is a graduate-level experimental methods class taught by MCF.
In previous years, students have conducted replication projects [@boyce2023; @hawkins2018].
In Fall 2023, students were offered the option to complete a "rescue" project where they re-replicated one of the unsuccessful replications from a previous year.
Students could also opt to do a normal replication instead.
If they opted for a rescue, they were given the option to participate in a collaborative meta-science project and earn authorship on the current paper based on 1) completing a rescue project, 2) reviewing another student's rescue project for reproducibility, and 3) reviewing and approving the final manuscript.
We report on the result of the `r num_proj` rescue projects by students who completed data collection and opted to be a part of the paper.

@@

Students were required to think of reasons the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions was given to students (<https://osf.io/dfwk2>).
In general, we encouraged students to add manipulation checks as appropriate and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the Cognitive Reflection Test (CRT) which many online participants are over-familiar with to using the newer CRT-2 which is used less but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map.
The replication ported this experiment online and had participants click a drop down menu to select left or right for each turn.
The rescue removed the unwieldy drop down menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs (VB and BP), students pre-registered and ran their samples.

With one exception, samples were collected on Prolific [the rescue of @yeshurun2003 ran in-person on the Stanford student subject pool].
We tried to power studies adequately [with a target of 2.5x original following @simonsohn2015], but due to cost constraints, not all studies were powered at this level.
The rescues had on average `r round(m,2)` times the original sample post-exclusion (median: `r round(mid,2)`, IQR: `r round(low,2)` - `r round(high,2)`, minimum: `r round(lowest, 2)`, maximum: `r round(highest,2)`, see Table \@ref(tab:tab-size) for all sample sizes).
Across the `r num_proj-1` Prolific studies, we spent \$`r pluck(cost,1,1) |> round()`, for an average of \$`r (pluck(cost,1,1)/(num_proj-1)) |> round()` per project.

@@

As a complement to the subjective rating of overall success, we statistically compared one key measure of interest for each study, following @boyce2023.
In order to statistically compare the key measures, we needed effect sizes reported in the same way for each original study, first replication, and rescue.
When effects were not reported in consistent ways across original and replications, we recalculated effects from raw data when necessary to obtain comparable values.

@@

We recorded the same set of potential correlates that were used in @boyce2023 for original, first replication, and rescue (these were already rated for original and first replication).
These potential correlates included the subfield of the study (cognitive, social, or other psychology), its publication year, experimental design features including whether it was a within- or between-subject design, whether each condition was instantiated with one vignette or multiple, how many items each participant saw, and whether there were open materials and open data.

For the original study and each replication, we recorded the number of participants after exclusions.
For studies where extra conditions were dropped in some replications, we counted only the participants in the key conditions all replications had for comparability.
For instance, if an original study compared between two critical conditions but also had a baseline control, we did not count the participants in the baseline condition if a replication did not include this condition.
We also recorded whether each study was conducted online using a crowd-sourced platform or not.

@@

Scientists' intuitions about whether a replication is successful and whether an effect provides support for a hypothesis (including heuristic cutoffs like p\<.05) do not always align with measures of statistical consistency [@patil2016].
For instance, two studies may both find that condition 1 results in a significantly higher outcome measure than condition 2, but the effect magnitudes may be sufficiently different that they are statistically unlikely to have come from the same population.
On the other hand, one study may find a statistically significant, but imprecisely estimated effect in a small sample, and a second study may find a near-zero (null) effect, but the effect estimates from the two studies may be statistically compatible, despite one supporting a hypothesized difference and the other not.

@@

library(kableExtra)
knitr::kable(cors, digits = 3, align = "rcc", format = "latex", booktabs = T, linesep = "", caption = "Correlations between an individual predictor and the subjective replication score of the rescue project. The first set of predictors were pre-registered based on the correlates used in Boyce et al. (2023). Original authors at Stanford refers to at the time of the first replication as this may mean the first replication was not independent. For details about how the predictors were coded, see the Methods section of Boyce et al. (2023). The last three predictors were added post-hoc.") |> row_spec(11, hline_after = T)
```

@@

The strongest correlates of rescue success were open materials, a small sample size on the first replication, a small sample size on the first replication relative to the original sample size, and a large rescue sample size relative to the first replication.
None of the pre-registered correlates meet the conventional significance threshold; the two correlates based on ratios that reflect a relatively smaller replication sample size are marginally significant.

Small replication samples relative to original and rescue could be due to both a) powering a replication according to a reported large effect size or b) difficulties with recruitment or high exclusion rates leading to a smaller than intended sample.
Since relative sizes of the studies may play a role in replication success and how probative replications are, we show the sample sizes in Table \@ref(tab:tab-size).

@@

@child2018 studied reading times for passages of text with different emotional valences written from either a third-person or second-person perspective.
They found a significant interaction between perspective and valence: participants read positively-valenced passages faster than negatively-valenced passages, but only when they were written from a second-person perspective.
The first replication used a similar sample size to the original study (40) and found an effect that trended in the same direction, but was substantially smaller.
The rescuer identified several possible reasons for this failure, foremost among them that the replication was conducted online while the original was conducted in person.
It is reasonable to expect online studies of reading time to be noisier than in-person studies because the environments participants completed the study in varied much more online.
The rescue attempt was still online, but used a much larger sample size to account for noisier data.
Still, the rescue found a non-significant effect that trended in the opposite direction from the original and first replication.
Even though the first replication trended in the same direction and there was an identifiable reason for the smaller effect size, increasing sample size did not rescue the effect.

@@

One potentially salient limitation of our work here is that the rescue projects (and the first replications) were primarily conducted by first-year PhD students; hence the generality of our conclusions to other investigators is unknown. We believe, however, that this is more of an apparent than an actual limitation. Students were enrolled in a highly competitive graduate program where it is expected that PhD students author peer-reviewed papers; several of the rescuers already had published first-authored peer reviewed articles. Thus, while we refer to the rescuers as students, they are investigators doing experiments that become part of the literature.

@@

In many cases, researchers may have strong prior beliefs about where a replication will succeed, and these priors may influence a researchers interpretation of a failed replication, and thus their calculations about where a re-replication will succeed or is worth attempting. For instance, a researcher who has high credence in the main result is more likely to attribute failure to specific details in the replication, and thus may want to do a rescue where those details are adjusted. On the other hand, a researcher who is dubious of the methods, results, or statistical approaches in an experiment, or who has previously failed to extend the experimental result, may be quicker to interpret a non-replication as a more definitive negative result. What a researcher chooses to do may also depend on the perceived strength of the replication, what size effects they think are relevant, and how important they think the target effect is. Thus, in some cases, when to re-replicate may be a question of prior beliefs and perceived utility rather than just likelihood of a re-replication succeeding.

>>>>

In psychology, large-scale replication projects have found that around half of studies successfully replicate.
Across 100 studies with positive results, the reproducibility project in psychology (RP:P) replicated 36%-47% depending on the metric for replication success [@opensciencecollaboration].
With large multi-site samples, Many Labs 1 replicated 11/13 effects (84.6%), Many Labs 2 replicated 14/28 effects (50%), and Many Labs 3 replicated 3/10 effects (30%) [@klein2014; @klein2018; @ebersole2016].
@camerer2018 examined 21 behavioral social science studies and successfully replicated 12-14 of them (57%-67%) depending on the metric used.
@boyce2023 reported an average subjective replication score of 49% for 176 replications primary in psychology.

Psychology is not the only discipline where many studies do not replicate.
Large-scale replication projects in other disciplines have found 39/97 of studies with positive effects (40%) replicating in cancer biology [@errington2021], 11/18 studies (61%) replicating in economics [@camerer2016], and 31/40 studies (78%) replicating in experimental philosophy [@cova2021].
Across these disciplines, and especially in psychology, large-scale replications indicate that the probability of scientists failing to replicate published findings is substantial.

@@

Other reasons for replication failure are primarily due to the design of the replication study.
A replication could observe different results from the original due to methodological differences.
These differences include intentional adaptations, unavoidable changes from lack of access to the original instructions or materials, or changes that the replicators did not expect to be critical.
A difference in the details of recruitment, such as changing from recruiting participants in person to online, might mean that particular details of a study implementation are no longer appropriate, or the data quality controls are inadequate.
If there are differences in time, place, or subject population between the two studies, researchers may need to change the materials, instructions, or procedure to adapt the experimental paradigm to the new context.
Such adaptation creates more challenges in interpretation: too many changes could cause differing results, but so could too few changes.
In sum, when replications fail, we generally do not know why they failed, although we may speculate.

@@

While it is possible to speculate after the fact about potential causes of a single replication failure, in most circumstances these causes are hard to discern (absent clear evidence of errors).
Re-replication can serve as a way to test explanations for replication failure.
Re-replications can make modifications to the first replication and give another chance for researchers to measure an effect of interest.
But what are the chances that a re-replication succeeds?

@@

While @ebersole2020 looked at re-replication success rate in a high-resource setting with access to large, multi-site samples and expert input, most early-career scientists do not have comparable resources to devote to re-replication attempts.
On the other hand, early-career researchers may also be prone to a different distribution of causes of initial replication failure.
Thus, it is an open question what the re-replication rate is for failed replications by early-career researchers.
Further, estimating this rate could inform researchers' decisions on what to do next when a replication fails.

@@

PSYCH 251 is a graduate-level experimental methods class taught by MCF.
In previous years, students conducted replication projects [@boyce2023; @hawkins2018].
In Fall 2023, students were offered the option to complete a "rescue" project where they re-replicated one of the unsuccessful replications from a previous year.
Students could also opt to do a normal replication instead.
If they opted for a rescue, they were given the option to participate in a collaborative meta-science project and earn authorship on the current paper based on 1) completing a rescue project, 2) reviewing another student's rescue project for reproducibility, and 3) reviewing and approving the final manuscript.
We report on the result of the `r num_proj` rescue projects by students who completed data collection and opted to be a part of the paper.

@@

Students were required to think of reasons why the first replication might not have worked, and address them if they could.
A list of possible reasons and solutions was given to students (<https://osf.io/dfwk2>).
In general, we encouraged students to add manipulation checks as appropriate and to adapt their materials to online studies, which may benefit from attention checks and/or shorter duration.
For instance, the rescue of @paxton2012 switched from using the Cognitive Reflection Test (CRT) which many online participants are over-familiar with to using the newer CRT-2 which is used less but attempts to measure the same construct [@thomson2016].
The rescue of @jara-ettinger2022 discovered that the replication had accidentally used an illustrated version of the stimuli rather than the photographic stimuli used in the original and used the photographic stimuli instead.
@tarampi2016 had participants do a timed navigation task where they wrote L or R to indicate which way to turn at each intersection on a paper map.
The replication ported this experiment online and had participants use a dropdown menu to select left or right for each turn.
The rescue removed the unwieldy dropdown menus and opted to have participants press keys on the keyboard to indicate the direction of a turn, which seemed like a more natural interface.
Once experimental designs and analytic plans were approved by class TAs (VB and BP), students pre-registered and ran their samples.

With one exception, samples were collected on Prolific [the rescue of @yeshurun2003 ran in person on the Stanford student subject pool].
We tried to power studies adequately [with a target sample size of 2.5x the original study following @simonsohn2015], but due to cost constraints, not all studies were powered at this level.
The rescues had on average `r round(m,2)` times the original sample post-exclusion (median: `r round(mid,2)`, IQR: `r round(low,2)` - `r round(high,2)`, minimum: `r round(lowest, 2)`, maximum: `r round(highest,2)`, see Table \@ref(tab:tab-size) for all sample sizes).
Across the `r num_proj-1` Prolific studies, we spent \$`r pluck(cost,1,1) |> round()`, for an average of \$`r (pluck(cost,1,1)/(num_proj-1)) |> round()` per project.

@@

As a complement to the subjective rating of overall success, we statistically compared one key measure of interest for each study, following @boyce2023.
In order to statistically compare the key measures, we needed effect sizes reported in the same way for each original study, first replication, and rescue.
When effects were not reported in consistent ways across original and replications, we recalculated effects from raw data to obtain comparable values.

@@

We recorded the same set of potential correlates that were used in @boyce2023 for original, first replication, and rescue (these were already rated for original and first replication).
These potential correlates included the subfield of the study (cognitive, social, or other psychology), its publication year, experimental design features including whether it used a within- or between-subject design, whether each condition was instantiated with one vignette or multiple, how many items each participant saw, and whether there were open materials and open data.

For the original study and each replication, we recorded the number of participants after exclusions.
For studies where extra conditions were dropped in some replications, we counted only the participants in the key conditions all replications had for comparability.
For instance, if an original study compared between two critical conditions but also had a baseline control, we did not count the participants in the baseline condition if a replication did not include this condition.
We also recorded whether each study was conducted using an online crowdsourcing platform or not.

@@

Scientists' intuitions about whether a replication is successful and whether an effect provides support for a hypothesis (including heuristic cutoffs like p\<.05) do not always align with measures of statistical consistency [@patil2016].
For instance, two studies may both find that condition 1 results in a significantly higher outcome measure than condition 2, but the effect magnitudes may be sufficiently different that they are statistically unlikely to have come from the same population.
On the other hand, one study may find a statistically significant, but imprecisely estimated, effect in a small sample, and a second study may find a near-zero (null) effect. The effect estimates from the two studies might be statistically compatible, despite one supporting a hypothesized difference and the other not.

@@

library(kableExtra)
knitr::kable(cors, digits = 3, align = "rcc", format = "latex", booktabs = T, linesep = "", caption = "Correlations between individual predictors and the subjective replication score of the rescue project. The first set of predictors were pre-registered based on the correlates used in Boyce et al. (2023). Original authors at Stanford refers to whether the any of the study's authors were affiliated with Stanford at the time of the first replication, as this may mean the first replication was not independent. For details about how the predictors were coded, see the Methods section of Boyce et al. (2023). The last three predictors were added post-hoc.") |> row_spec(11, hline_after = T)
```

@@

The strongest correlates of rescue success were open materials, a small sample size in the first replication, a small sample size in the first replication relative to the original sample size, and a large rescue sample size relative to the first replication.
None of the pre-registered correlates meet the conventional significance threshold; the two correlates based on ratios that reflect a relatively smaller replication sample size are marginally significant.

Small replication samples relative to original and rescue could be due to both a) powering a replication according to a reported large effect size or b) difficulties with recruitment or high exclusion rates leading to a sample size smaller than intended.
Since relative sizes of the studies may play a role in replication success and how probative replications are, we show the sample sizes in Table \@ref(tab:tab-size).

@@

@child2018 studied reading times for passages of text with different emotional valences written from either a third-person or second-person perspective.
They found a significant interaction between perspective and valence: participants read positively-valenced passages faster than negatively-valenced passages, but only when they were written from a second-person perspective.
The first replication used a similar sample size to the original study (40) and found an effect that trended in the same direction, but was substantially smaller.
The rescuer identified several possible reasons for this failure, foremost among them that the replication was conducted online while the original was conducted in person.
It is reasonable to expect online studies of reading time to be noisier than in-person studies because the environments participants completed the study in vary much more online.
The rescue attempt was still online, but used a much larger sample size to account for noisier data.
Still, the rescue found a non-significant effect that trended in the opposite direction from the original and first replication.
Even though the first replication trended in the same direction and there was an identifiable reason for the smaller effect size, increasing sample size did not rescue the effect.

@@

One potentially salient limitation of our work here is that the rescue projects (and the first replications) were primarily conducted by first-year PhD students; hence the generality of our conclusions to other investigators is unknown. We believe, however, that this is more of an apparent than an actual limitation. Students were enrolled in a highly competitive graduate program where it is expected that PhD students author peer-reviewed papers. Several of the rescuers had already published first-authored peer reviewed articles. Thus, while we refer to the rescuers as students, they are investigators doing experiments that become part of the literature.

@@

In many cases, researchers may have strong prior beliefs about where a replication will succeed, and these priors may influence a researcher's interpretation of a failed replication, and thus their calculations about where a re-replication will succeed or is worth attempting. For instance, a researcher who has high credence in the main result is more likely to attribute failure to specific details in the replication, and thus may want to do a rescue where those details are adjusted. On the other hand, a researcher who is dubious of the methods, results, or statistical approaches in an experiment, or who has previously failed to extend the experimental result, may be quicker to interpret a non-replication as a more definitive negative result. What a researcher chooses to do may also depend on the perceived strength of the replication, what size effects they think are relevant, and how important they think the target effect is. Thus, in some cases, when to re-replicate may be a question of prior beliefs and perceived utility rather than just likelihood of a re-replication succeeding.
